Healthcare Insights and Sociodemographic Analysis of New York: A Comprehensive Analysis of Patient Records in 2013

### 2023-10-01 STAT405Project.qmd
basically a lot of shit

*(last updated 09/23/2023)*

@@

@@

```{r}
#| echo: false
#install.packages("readr")
#install.packages("xtable")
library(readr)
library(xtable)
options(xtable.floating = FALSE)
options(xtable.timestamp = "")
data_set <- read_csv("Hospital_Inpatient_Discharges__SPARCS_De-Identified___2013.csv",show_col_types = FALSE)

@@

```{r, results='asis'}
#| echo: false
length_of_stay <- strtoi(cleaned_data$`Length of Stay`)
plot(length_of_stay, y,  xlab = "Length of Stay", main = "Price vs Length of Stay", ylab = "Price")

lmfs <- lm(y ~ length_of_stay)
xtable(lmfs, comment = FALSE)
abline(lmfs, col = "RED")
```

@@

```{r}
#| echo: false
cleaned_data$'Total Costs'[cleaned_data$'Total Costs' > quantile(cleaned_data$'Total Costs', 0.95)] <- quantile(cleaned_data$'Total Costs', 0.95)

summary(cleaned_data$'Total Costs')

boxplot(cleaned_data$'Total Costs', 
        main="Box Plot of Total Costs (Outliers Truncated)",
        ylab="Total Costs",
        horizontal = TRUE)
```

@@

```{r, warning = FALSE, fig.height = 6, fig.width = 50}
#| echo: false
options(repr.plot.width = 100, repr.plot.height = 100)

cleaned_data$`Payment Typology 1` <- as.factor(cleaned_data$`Payment Typology 1`)
payments <- table(cleaned_data$`Payment Typology 1`)
pie(payments, main = "Payment Type Distribution", col=rainbow(length(payments)), labels = NA, )
legend(1,1,1, legend = levels(cleaned_data$`Payment Typology 1`), fill = rainbow(length(payments)), cex = 0.7)

@@

grid()
```

@@

```{r}
#| echo: false
ageGroups <- table(cleaned_data$`Age Group`)
barplot(ageGroups, main = "Age Group Distribution", xlab = "Age Group", 
        ylab = "Frequency", col = "green", border = "black", ylim = c(0, 300000)) 
```

@@

heatmap(heatmap_data,
        Rowv = NA,          
        Colv = NA,       
        col = colorRampPalette(c("white", "blue"))(50),  
        xlab = "Gender",    
        main = "Gender v. Age Group",  
        labRow = row_names, 
        labCol = col_names)

@@

```{r}
#| echo: false
#install.packages("ggplot2")
library(ggplot2)
ggplot(cleaned_data, aes(x = Gender, fill = Race)) +
  geom_bar() +
  labs(
    title = "Stacked Bar Chart by Gender and Race",
    x = "Gender",
    y = "Count"
  ) +
  scale_fill_manual(
    values = c("White" = "lightblue", "Black" = "lightcoral", "Other" = "lightgreen"),
    name = "Race"
  ) +
  theme_minimal()

@@

```{r warning = FALSE}
#| echo: false
#install.packages("ggplot2")
ggplot(cleaned_data) + 
  aes(x = cleaned_data$'APR Risk of Mortality', y = cleaned_data$"Total Costs", color = "salmon") + 
  geom_boxplot() +
  labs(title = "Total Costs on Risk of Mortality",
       x = "APR Risk of Mortality",  # X-axis title
       y = "Total Costs")  # Y-axis title

## Plot 9: Boxplots on the Total Costs on the Risk of Mortality

## Plot 10:

## Plot 11:

## Plot 12:

>>>>

*(last updated 09/30/2023)*

@@

\newpage

@@

```{r, include = FALSE}
#| echo: false
# install.packages("readr")
# install.packages("xtable")
library(readr)
library(dplyr)
library(xtable)
options(xtable.floating = FALSE)
options(xtable.timestamp = "")
data_set <- read_csv("Hospital_Inpatient_Discharges__SPARCS_De-Identified___2013.csv", show_col_types = FALSE)

@@

```{r, results='asis'}
#| echo: false
length_of_stay <- strtoi(cleaned_data$`Length of Stay`)

# Set scipen to a large value to avoid scientific notation
options(scipen = 10)

# Create a scatter plot
plot(length_of_stay, y, xlab = "Length of Stay", main = "Price vs Length of Stay", ylab = "Price")

# Reset scipen to its default value after the plot
options(scipen = 0)

lmfs <- lm(y ~ length_of_stay)
abline(lmfs, col = "RED")

```
<span style="font-size: 14px; font-style: italic;">Figure 1a: Length of Stay on Price Regression Dotplot</span>

```{r, results='asis'}
#| echo: false
table_lmfs <- xtable::xtable(lmfs, comment = NA, caption = " Figure 1.b: Linear Model Summary")

cat("\\begin{center}\n")
print(table_lmfs, comment = FALSE)
cat("\\end{center}\n")
```
<i>Figure 1b: Table of Linear Regression Modeling Data</i>

@@

```{r}
#| echo: false
cleaned_data$"Total Costs"[cleaned_data$"Total Costs" > quantile(cleaned_data$"Total Costs", 0.95)] <- quantile(cleaned_data$"Total Costs", 0.95)

#summary(cleaned_data$"Total Costs")

boxplot(cleaned_data$"Total Costs",
    main = "Box Plot of Total Costs (Outliers Truncated)",
    ylab = "Total Costs",
    horizontal = TRUE
)

```
<i>Figure 2: Total Costs Boxplot</i>

@@

```{r, warning = FALSE, fig.height = 6, fig.width = 50}
#| echo: false
# Set the margin to include more space at the bottom
par(mar = c(5, 4, 4, 2) + 0.1)

cleaned_data$`Payment Typology 1` <- as.factor(cleaned_data$`Payment Typology 1`)
payments <- table(cleaned_data$`Payment Typology 1`)
pie(payments, main = "Payment Type Distribution", col = rainbow(length(payments)), labels = NA)
legend(1, 1, 1, legend = levels(cleaned_data$`Payment Typology 1`), fill = rainbow(length(payments)), cex = 0.7)

```
<i>Figure 3: Pie Chart of Payment Distribution</i>

@@

grid()
```
<span style="font-size: 14px; font-style: italic;">Figure 4: Trends of Total Charges and Legnth of Stay </span>

@@

```{r}
#| echo: false
ageGroups <- table(cleaned_data$`Age Group`)
barplot(ageGroups,
    main = "Age Group Distribution", xlab = "Age Group",
    ylab = "Frequency", col = "green", border = "black", ylim = c(0, 300000)
)
mtext("Figure 5: Age Group and Distribtion", side = 1, line = 4, adj = -0.1, cex = 0.7, font = 3, col = "black")

```
<span style="font-size: 14px; font-style: italic;">Figure 5: Age Group Distribution of Patients in New York</span>

@@

heatmap(heatmap_data,
    Rowv = NA,
    Colv = NA,
    col = colorRampPalette(c("white", "blue"))(50),
    xlab = "Gender",
    main = "Gender v. Age Group",
    labRow = row_names,
    labCol = col_names
)
mtext("Figure 6: Gender vs. Age Group Heatmap", side = 1, line = 4, adj = -0.1, cex = 0.7, font = 3, col = "black")

```
<span style="font-size: 14px; font-style: italic;">Figure 6: Gender vs Age Group Heatmapt</span>

@@

```{r}
#| echo: false
# install.packages("ggplot2")
library(ggplot2)

plot7 <- ggplot(cleaned_data, aes(x = Gender, fill = Race)) +
  geom_bar() +
  labs(
    title = "Stacked Bar Chart by Gender and Race",
    x = "Gender",
    y = "Count"
  ) +
  scale_fill_manual(
    values = c("White" = "lightblue", "African American" = "lightcoral", "Other" = "lightgreen"),
    breaks = c("White", "Black", "Other"),
    labels = c("White", "Black", "Other"),
    name = "Race"
  ) +
  theme_minimal()

plot7 + ggtitle("Data Source: mtcars dataset, ggplot2 package")

```
<span style="font-size: 14px; font-style: italic;">Figure 7: Stacked Bar Chart by Race and Gender</span>

## Plot Description and Inferencing

In the analysis of patient demographics in the year 2013, a bar chart was employed to visually represent the breakdown of patients by gender and further stratify each gender by race. A notable observation from the chart is the significant predominance of female patients over male patients during this period.

Upon closer examination of the male demographic, the distribution reveals a relatively balanced representation between individuals identified as white and those classified as not white. This suggests a relatively even distribution among male patients concerning racial categories. In contrast, the female demographic exhibits a slightly more diverse distribution across racial categories. This observation implies that, among female patients, there is a greater degree of heterogeneity in terms of racial composition compared to their male counterparts.

These findings contribute valuable insights into the gender and racial composition of the patient population in 2013, emphasizing the need for a nuanced understanding of demographic trends within healthcare settings. Further exploration and analysis of these demographic patterns may reveal additional insights relevant to healthcare planning, resource allocation, and the provision of patient-centered care.

\newpage

@@

```{r warning = FALSE}
#| echo: false
# install.packages("ggplot2")
ggplot(cleaned_data) +
    aes(x = cleaned_data$"APR Risk of Mortality", y = cleaned_data$"Total Costs", color = "salmon") +
    geom_boxplot() +
    labs(
        title = "Total Costs on Risk of Mortality",
        x = "APR Risk of Mortality", # X-axis title
        y = "Total Costs"
    ) # Y-axis title

```
<span style="font-size: 14px; font-style: italic;">Figure 8: Boxplots on Total Cost and Mortality</span>

## Plot Description and Inferencing
In Figure 8, our objective was to assess the central tendency of total hospital costs across different levels of Risk of Mortality. Employing separate box plots for each severity category allowed us to discern distinct patterns in the distribution of total costs associated with varying degrees of mortality risk. Notably, when focusing on Extreme cases of mortality, a clear trend emerged—median total costs were significantly higher, accompanied by a broader spread in the distribution. This observation implies that patients facing an Extreme risk of mortality not only incur higher median costs but also exhibit greater variability in the financial implications of their healthcare, suggesting a more diverse range of medical interventions and resource utilization.

This finding aligns with our initial expectation that higher levels of risk of mortality correspond to elevated costs of care. The discernible trend in the box plots reinforces the intuitive understanding that patients with a greater severity of illness often require more extensive and resource-intensive medical interventions, leading to increased healthcare costs. Beyond its immediate implications, this nuanced analysis offers valuable insights for healthcare planning and resource allocation, emphasizing the need for tailored strategies based on the specific needs associated with varying risk levels. This approach contributes to more effective and efficient healthcare delivery, promoting a patient-centered and economically sustainable healthcare system.

\newpage

## Plot 9: Violin Plot Comparing New York City and Western New York's Severity of Illness

```{r warning = FALSE}
#| echo: false
subsetted_data <- cleaned_data[cleaned_data$`Health Service Area` %in% c("New York City", "Western NY"), ]

# Load the ggplot2 library
library(ggplot2)

# Create a violin plot for APR Severity of Illness Code
ggplot(subsetted_data, aes(x = factor(`Health Service Area`), y = `APR Severity of Illness Code`, fill = `Health Service Area`)) +
  geom_violin() +
  labs(
    x = "Health Service Area",
    y = "APR Severity of Illness Code",
    fill = "Health Service Area",
    title = "Distribution of APR Severity of Illness Code in New York City"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
<span style="font-size: 14px; font-style: italic;">Figure 9: NYC and Western NY Severity of Illness Comparsion </span>

## Plot Description and Inferencing
Figure 9 delves into a detailed breakdown of hospital patients, distinguishing between those originating from New York City (NYC) and Western New York (WNY). The focus is on the distribution of patients' APR severity codes, measured on a 1 to 4 scale (1 being mild, 2 being moderate, 3 being serious, and 4 being extreme). Notably, the violin plot reveals a higher frequency of APR severity code 2 assignments for patients from NYC compared to WNY. This divergence prompts exploration into potential variations in labeling practices between the two regions. However, beyond this distinction, the overall shapes of the violin distributions appear similar, suggesting a general alignment in severity profiles. This observation implies a degree of consistency in patient conditions based on APR severity codes, despite the regional differences in labeling practices. Understanding such variations is crucial for refining medical coding precision and ensuring uniformity in patient severity assessments across diverse geographical contexts.

\newpage

## Plot 10: Distribution of APR Severity of Illness for Medicaid Patients

```{r, warning = FALSE}
#| echo: false
# Load the required libraries
library(ggplot2)

medicaid_data <- cleaned_data %>%
  filter(`Payment Typology 1` == "Medicaid")

# Create a ridgeline plot to visualize APR Severity of Illness
ggplot(medicaid_data, aes(x = `APR Severity of Illness Code`, y = ..density.., fill = `APR Severity of Illness Code`)) +
  geom_density(alpha = 0.6) +
  labs(
    x = "APR Severity of Illness Code",
    y = "Density",
    fill = "APR Severity of Illness Code",
    title = "Distribution of APR Severity of Illness for Medicaid Patients"
  ) +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1")

```
<span style="font-size: 14px; font-style: italic;">Figure 10: APR Severity of Illness for Medicaid Patients </span>

## Plot Description and Inferencing
In our examination of Figure 10, our focus shifted toward a specific subset of patients—those under Medicaid care. Medicaid, a health program designed to support individuals classified as economically disadvantaged, prompted our investigation into the severity of diagnoses within this demographic. Our objective was to discern whether patients under Medicaid receive comparable attention and care in contrast to their counterparts outside of the program.

Distinct from the broader population, our analysis revealed a notable prevalence of patients with "mild" severity codes among those under Medicaid care. This intriguing observation necessitates careful consideration of various contributing factors. While we refrain from making definitive conclusions about the intentional prioritization of severity levels for Medicaid patients, this finding presents a compelling avenue for further exploration.

The presence of a higher number of "mild" severity cases within the Medicaid subset raises questions about potential disparities in healthcare provision. It is important to acknowledge that a comprehensive understanding of this phenomenon requires additional contextual data. Future investigations could delve deeper into the intricate dynamics surrounding patient severity levels, ensuring a nuanced exploration that considers various influencing factors.

\newpage

## Plot 11: Facet Grid of Dotplots for Total Charges on Length of Stay

```{r warning = FALSE}
#| echo: false
library(ggplot2)

ggplot(cleaned_data, aes(x = cleaned_data$`Length of Stay`, y = cleaned_data$`Total Charges`)) +
  geom_point(size = 0.2) +
  facet_grid(Race ~ `Type of Admission`) +
  labs(title = "Length of Stay v. Total Charges", x = "Length of Stay", y = "Total Charges") +
  scale_y_continuous(labels = scales::comma, 
                     breaks = scales::pretty_breaks(), 
                     limits = c(0, max(cleaned_data$`Total Charges`)), # 
                     expand = c(0, 0)) +
  theme(axis.text = element_text(size = 8))

```
<span style="font-size: 14px; font-style: italic;">Figure 11: Dotplots for Total Charges on Length of Stay </span>

## Plot Description and Inferencing

In Figure 11, our aim was to dissect assessments based on both race and the type of admission. Through our analysis, a discernible pattern emerged, revealing consistent spikes in each category of total charges, particularly towards the lower end of the length of stay spectrum. This observation suggests that, irrespective of race, there are commonalities in the distribution of total charges, with pronounced concentration points occurring for shorter stays.

Upon closer examination, the data points appear to be uniformly distributed across all racial categories, implying a degree of similarity in the cost distribution patterns among different racial groups. Interestingly, the primary differentiating factor contributing to variations in total costs is identified as the type of admission. This finding underscores the significance of admission characteristics in influencing healthcare costs, emphasizing that differences in total charges are more prominently linked to variations in admission types rather than racial distinctions.

These insights shed light on the nuanced interplay between race, type of admission, and healthcare costs. Understanding these relationships is crucial for healthcare administrators and policymakers in designing targeted strategies for cost-effective care delivery, ensuring equitable access, and fostering a healthcare system that is both inclusive and economically sustainable.

\newpage

## Plot 12: Facet Density Plots of Total Cost per Age-Group Buckets

```{r warning = FALSE}
#| echo: false
ggplot(cleaned_data, aes(x = `Total Costs`)) +
  geom_density(fill = "blue", alpha = 0.5) +
  facet_grid(`Age Group` ~ .) +
labs(title = "Density Plot of Total Costs per Age-Group Bucket", x = "Total Costs", y = "Density");

```
<span style="font-size: 14px; font-style: italic;">Figure 12: Density Plots of Total Costs per Age-Groups</span>

## Plot Description and Inferencing
In Figure 12, we analyzed age-specific total costs by charting density plots using a facet grid. For patients under 17, costs skewed right, suggesting lower expenses, potentially due to the overall good health of children. In contrast, as age increases, a broader and more dispersed peak in total costs emerges, indicating that older patients require more extensive and potentially costlier healthcare interventions, such as hospice services. This exploration sheds light on age-related patterns in healthcare costs, guiding strategies for resource allocation and care planning across different life stages.

\newpage

### References
New York State Department of Health. "Hospital Inpatient Discharges (SPARCS De-Identified)." data.ny.gov, 2013, https://health.data.ny.gov/Health/Hospital-Inpatient-Discharges-SPARCS-De-Identified/npsr-cm47.
